---
title: "Binomial Regression for Survival and Competing Risks Data"
author: Klaus Holst & Thomas Scheike
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    fig_width: 7.15
    fig_height: 5.5 
vignette: >
  %\VignetteIndexEntry{Binomial Regression for Survival and Competing Risks Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mets)
```

Binomial Regression for censored data 
=====================================

Details
=======

The binreg function does direct binomial regression for one time-point, $t$, fitting the model 
\begin{align*}
P(T \leq t, \epsilon=1 | X )  & = \mbox{expit}( X^T \beta)  = F_1(t,X,\beta)
\end{align*}
to an IPCW adjusted estmating equation (EE) with response  $Y(t)=I(T \leq t, \epsilon=1 )$
\begin{align*}
 U(\beta,\hat G_c) = &  X ( Y(t) \frac{ \Delta(t) }{\hat G_c(T_i \wedge t)} - \mbox{expit}( X^T \beta)) = 0,
\end{align*}
with $G_c(t)=P(C>t)$, the  censoring survival distribution, and  with $\Delta(t) = I( C_i > T_i \wedge t)$
the indicator of being uncensored at time $t$ (type="I"). 
The default type="II" is to augment with a censoring term, that is solve 
\begin{align*}
  &  U(\beta,\hat G_c) - \int_0^t X \frac{\hat E(Y(t)| T>u)}{\hat G_c(u)} d\hat M_c(u) =0 
\end{align*}
where $M_c(u)$ is the censoring martingale, this typically improves the performance. This is equivlent to the 
pseudo-value approach (see Overgaard (2025)).

The influence function for the type="II" estimator is 
\begin{align*}
    U(\beta,G_c) - \int_0^t X \frac{E(Y| T>u)}{G_c(u)} d M_c(u)  
    + \int_0^t  \frac{E(X| T>u) E(Y| T>u)}{G_c(u)} d M_c(u) + \int_0^t \frac{E( X Y| T>u)}{G_c(u)} d M_c(u)  
\end{align*}
and for type="I"
\begin{align*}
  &  U(\beta) + \int_0^t \frac{E( X Y| T>u)}{G_c(u)} d M_c(u).
\end{align*}
The means $E(X Y(t) | T>u)$ and $E(Y(t)| T>u)$ are estimated by IPCW estimators among survivors to get 
estimates of the influence functions.

The function logitIPCW instead considers 
\begin{align*}
 U^{glm}(\beta,\hat G_c) = & \frac{ \Delta(t) }{\hat G_c(T_i \wedge t)}  X  ( Y(t) - \mbox{expit}( X^T \beta)) = 0.
\end{align*}
This score equation is quite similar to those of the binreg, and exactly the same when the censoring model is fully-nonparametric.

The logitIPCW has influence function 
\begin{align*}
  &  U^{glm}(\beta,G_c) + \int_0^t \frac{E( X ( Y - F_1(t,\beta)) | T>u)}{G_c(u)} d M_c(u)  
\end{align*}

Which estimator performs the best depends on the censoring distribution and it seems that the binreg with type="II" performs 
overall quite nicely (see Blanche et al (2023) and Overgaard (2024)).  For the full estimated censoring model all estimators have the
same influence function (see Blanche et al (2023)).

Additional functions logitATE, and binregATE computes the average treatment effect. 
We demonstrate this in another vignette. 

The functions logitATE/binregATE can be used there is no censoring and we thus have simple binary outcome. 

The variance is based on sandwich formula with IPCW adjustment (using the influence functions), 
and naive.var is the variance under known censoring model. The influence functions are stored in the output. 
Clusters can be specified to get cluster corrected standard errors. 

Examples
========


```{r}
 library(mets)
 options(warn=-1)
 set.seed(1000) # to control output in random noise just below.
 data(bmt)
 bmt$time <- bmt$time+runif(nrow(bmt))*0.01

 # logistic regresion with IPCW binomial regression 
 out <- binreg(Event(time,cause)~tcell+platelet,bmt,time=50)
 summary(out)
```

We can also compute predictions using the estimates 
```{r}
 predict(out,data.frame(tcell=c(0,1),platelet=c(1,1)),se=TRUE)
```

Further the censoring model can depend on strata 

```{r}
 outs <- binreg(Event(time,cause)~tcell+platelet,bmt,time=50,cens.model=~strata(tcell,platelet))
 summary(outs)
```

Absolute risk differences and ratio
===================================

Now for illustrations I wish to consider the absolute risk difference
depending on tcell 

```{r}
 outs <- binreg(Event(time,cause)~tcell,bmt,time=50,cens.model=~strata(tcell))
 summary(outs)
```

the risk difference is 

```{r}
ps <-  predict(outs,data.frame(tcell=c(0,1)),se=TRUE)
ps
sum( c(1,-1) * ps[,1])
```

Getting the standard errors are easy enough since the two-groups are
independent. In the case where we in addition had adjusted for other covariates, however, we would need 
the to apply the delta-theorem thus using the relevant covariances along the lines of 

```{r}
dd <- data.frame(tcell=c(0,1))
p <- predict(outs,dd)

riskdifratio <- function(p,contrast=c(1,-1)) {
   outs$coef <- p
   p <- predict(outs,dd)[,1]
   pd <- sum(contrast*p)
   r1 <- p[1]/p[2]
   r2 <- p[2]/p[1]
   return(c(pd,r1,r2))
}
     
estimate(outs,f=riskdifratio,dd,null=c(0,1,1))
```

same as 

```{r}
run <- 0
if (run==1) {
library(prodlim)
pl <- prodlim(Hist(time,cause)~tcell,bmt)
spl <- summary(pl,times=50,asMatrix=TRUE)
spl
}
```


Augmenting the Binomial Regression 
===================================

Rather than using a larger censoring model we can also compute an  augmentation term and
then fit the binomial regression model based on this augmentation term. 
Here we compute the augmentation based on stratified non-parametric estimates of $F_1(t,S(X))$, 
where $S(X)$ gives strata based on $X$ as a working model. 

 Computes  the augmentation term for each individual as well as the sum
\begin{align*}
 A & = \int_0^t H(u,X) \frac{1}{S^*(u,s)} \frac{1}{G_c(u)} dM_c(u)
\end{align*}
 with 
\begin{align*}
 H(u,X) & = F_1^*(t,S(X)) - F_1^*(u,S(X))
\end{align*}
using a KM for $G_c(t)$ and a working model for cumulative baseline
related to $F_1^*(t,s)$ and $s$ is strata, $S^*(t,s) = 1 - F_1^*(t,s) - F_2^*(t,s)$. 

Standard errors computed under assumption of correct but estimated $G_c(s)$ model.

```{r}
 data(bmt)
 dcut(bmt,breaks=2) <- ~age 
 out1<-BinAugmentCifstrata(Event(time,cause)~platelet+agecat.2+
			  strata(platelet,agecat.2),data=bmt,cause=1,time=40)
 summary(out1)

 out2<-BinAugmentCifstrata(Event(time,cause)~platelet+agecat.2+
     strata(platelet,agecat.2)+strataC(platelet),data=bmt,cause=1,time=40)
 summary(out2)
```


SessionInfo
============

```{r}
sessionInfo()
```

