---
title: "Discrete Interval Censored Survival Models"
author: Klaus Holst & Thomas Scheike
date: "`r Sys.Date()`"
output:
  knitr:::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Discrete Interval Censored Survival Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mets)
```

Discrete Inteval Censored survival times 
========================================

\begin{align*}
   \mbox{logit}(P(T >t | x)) & = \log(G(t)) + x^T \beta              \\
                P(T >t | x)  & =  \frac{1}{1 + G(t) exp( x^T \beta) }         
\end{align*}
 

Input are intervals given by $]t_l,t_r]$ where t_r can be infinity
for right-censored intervals. When truly discrete $]0,1]$ will be an
observation at 1, and $]j,j+1]$ will be an observation at j+1.

Likelihood is maximized:
\begin{align*}
    \prod_i  P(T_i >t_{il} | x) - P(T_i> t_{ir}| x).            
\end{align*}

This model is also called the cumulative odds model 
\begin{align*}
     P(T \leq t | x)  & =  \frac{ G(t) exp( x^T \beta) }{1 + G(t) exp( x^T \beta) }.
\end{align*}
and $\beta$ says something about the OR of probability of being before $t$. 

The baseline is parametrized as 
\begin{align*}
                G(t)  & =  \sum_{j \leq t} \exp( \alpha_j ) 
\end{align*}

An important consequence of the model is that for all cut-points $t$ we have the same OR parameters for the 
OR of being early or later than $t$. 

Discrete TTP 
=============


First we look at some time to pregnancy (discrete survival data) that is
right-censored, and set it up to fit the cumulative odds model by constructing the intervals appropriately: 

```{r}
library(mets)

data(ttpd) 
dtable(ttpd,~entry+time2)
out <- interval.logitsurv.discrete(Interval(entry,time2)~X1+X2+X3+X4,ttpd)
summary(out)
```

Now using this discrete survival model we simulate some data from this model 

```{r}
set.seed(1000) # to control output in simulatins for p-values below.
n <- 10000
Z <- matrix(rbinom(n*4,1,0.5),n,4)
outsim <- simlogitSurvd(out$coef,Z)
outsim <- transform(outsim,left=time,right=time+1)
outsim <- dtransform(outsim,right=Inf,status==0)
outss <- interval.logitsurv.discrete(Interval(left,right)~+X1+X2+X3+X4,outsim)
summary(outss)

pred <- predictlogitSurvd(out,se=TRUE)
plotSurvd(pred,se=TRUE)
```

And now taking some data and comparing with icenReg. We make the data fully interval censored/discrete by
letting also exact obsevations be discrete (thus being in an interval) by modifying the intervals. 

We consider the interval censored survival times for time from onset 
of diabetes to to diabetic nephronpathy, and make it discrete by 
fixing a nuber of fixed cutpoints.


```{r}
test <- 0 
if (test==1) {

require(icenReg)
data(IR_diabetes)
IRdia <- IR_diabetes
## removing fully observed data in continuous version, here making it a discrete observation 
IRdia <- dtransform(IRdia,left=left-1,left==right)
dtable(IRdia,~left+right,level=1)

ints <- with(IRdia,dInterval(left,right,cuts=c(0,5,10,20,30,40,Inf),show=TRUE) )
}
```


```{r}
if (test==1) {
ints$Ileft <- ints$left
ints$Iright <- ints$right
IRdia <- cbind(IRdia,data.frame(Ileft=ints$Ileft,Iright=ints$Iright))
dtable(IRdia,~Ileft+Iright)
# 
#       Iright   1   2   3   4   5 Inf
# Ileft                               
# 0             10   1  34  25   4   0
# 1              0  55  19  17   1   1
# 2              0   0 393  16   4   0
# 3              0   0   0 127   1   0
# 4              0   0   0   0  21   0
# 5              0   0   0   0   0   2

outss <- interval.logitsurv.discrete(Interval(Ileft,Iright)~+gender,IRdia)
#            Estimate Std.Err    2.5%    97.5%   P-value
# time1        -3.934  0.3316 -4.5842 -3.28418 1.846e-32
# time2        -2.042  0.1693 -2.3742 -1.71038 1.710e-33
# time3         1.443  0.1481  1.1530  1.73340 1.911e-22
# time4         3.545  0.2629  3.0295  4.06008 1.976e-41
# time5         6.067  0.7757  4.5470  7.58784 5.217e-15
# gendermale   -0.385  0.1691 -0.7165 -0.05351 2.283e-02
summary(outss)
outss$ploglik
# [1] -646.1946

fit <- ic_sp(cbind(Ileft, Iright) ~ gender, data = IRdia, model = "po")
# 
# Model:  Proportional Odds
# Dependency structure assumed: Independence
# Baseline:  semi-parametric 
# Call: ic_sp(formula = cbind(Ileft, Iright) ~ gender, data = IRdia, 
#     model = "po")
# 
#            Estimate Exp(Est)
# gendermale    0.385     1.47
# 
# final llk =  -646.1946 
# Iterations =  6 
# Bootstrap Samples =  0 
# WARNING: only  0  bootstrap samples used for standard errors. 
# Suggest using more bootstrap samples for inference
summary(fit)

## sometimes NR-algorithm needs modifications of stepsize to run 
## outss <- interval.logitsurv.discrete(Interval(Ileft,Iright)~+gender,IRdia,control=list(trace=TRUE,stepsize=1.0))
}

```


SessionInfo
============


```{r}
sessionInfo()
```


