---
title: "Mediation Analysis for survival data"
author: Klaus Holst & Thomas Scheike
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    fig_width: 7.15
    fig_height: 5.5        
vignette: >
  %\VignetteIndexEntry{Mediation Analysis for survival data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  #dev="png",
  comment = "#>"  
)
fullVignette <- Sys.getenv("_R_FULL_VIGNETTE_") %in% c("1","TRUE")
library(mets)
```

Overview 
========

Fit 

 * binomial-regression IPCW, binreg 
 * additive Lin-Ying model, aalenMets
 * cox model phreg 
 * standard logistic regression via binreg

in the context of mediation analysis using mediation weights as in the medFlex package. 

The mediator can 

 * binomial using glm-binomial.
 * multnomial via the mlogit function 

Both mediator and exposure must be coded as factors.

In the below example these are

 * mediator: gp.f
 * exposure : dnr.f

and the outcome model is concerned with the risk/hazard of cause=2. 

The key is that the standard errors are computed using the i.i.d influence
functions and a Taylor expansion to deal with the uncertainty from the
mediation weights. 


```{r}
 library(mets)
 runb <- 0
 options(warn=-1)
 set.seed(1000) # to control output in simulatins for p-values below.

n <- 200; k.boot <- 10; 

dat <- kumarsimRCT(n,rho1=0.5,rho2=0.5,rct=2,censpar=c(0,0,0,0),
          beta = c(-0.67, 0.59, 0.55, 0.25, 0.98, 0.18, 0.45, 0.31),
    treatmodel = c(-0.18, 0.56, 0.56, 0.54),restrict=1)
dfactor(dat) <- dnr.f~dnr
dfactor(dat) <- gp.f~gp
drename(dat) <- ttt24~"ttt24*"
dat$id <- 1:n
dat$ftime <- 1
```

Compute mediation weights   based on fitted  model 

```{r}
weightmodel <- fit <- glm(gp.f~dnr.f+preauto+ttt24,data=dat,family=binomial)
wdata <- medweight(fit,data=dat)

aaMss2 <- binreg(Event(time,status)~gp+dnr+preauto+ttt24+cluster(id),data=dat,time=50,cause=2)
summary(aaMss2)
aaMss22 <- binreg(Event(time,status)~dnr+preauto+ttt24+cluster(id),data=dat,time=50,cause=2)
summary(aaMss22)
```


```{r}
### binomial regression ###########################################################
aaMss <- binreg(Event(time,status)~dnr.f0+dnr.f1+preauto+ttt24+cluster(id),data=wdata,time=50,weights=wdata$weights,cause=2)
summary(aaMss)

ll <- mediatorSurv(aaMss,fit,data=dat,wdata=wdata)
summary(ll)
if (runb>0) { bll <- BootmediatorSurv(aaMss,fit,data=dat,k.boot=k.boot); summary(bll)}

### lin-ying model ################################################################
aaMss <- aalenMets(Surv(time/100,status==2)~dnr.f0+dnr.f1+preauto+ttt24+cluster(id),data=wdata,weights=wdata$weights)
ll <- mediatorSurv(aaMss,fit,data=dat,wdata=wdata)
summary(ll)
if (runb>0) { bll <- BootmediatorSurv(aaMss,fit,data=dat,k.boot=k.boot); summary(bll)}

### cox model ###############################################################################
aaMss <- phreg(Surv(time,status==2)~dnr.f0+dnr.f1+preauto+ttt24+cluster(id),data=wdata,weights=wdata$weights)
summary(aaMss)
ll <- mediatorSurv(aaMss,fit,data=dat,wdata=wdata)
summary(ll)
if (runb>0) { bll <- BootmediatorSurv(aaMss,fit,data=dat,k.boot=k.boot); summary(bll)}

### Fine-Gray #############################################################3
aaMss <- cifreg(Event(time,status)~dnr.f0+dnr.f1+preauto+ttt24+cluster(id),data=wdata,weights=wdata$weights,propodds=NULL,cause=2)
summary(aaMss)
ll <- mediatorSurv(aaMss,fit,data=dat,wdata=wdata)
summary(ll)
if (runb>0) { bll <- BootmediatorSurv(aaMss,fit,data=dat,k.boot=k.boot); summary(bll)}


### logit model  #############################################################3
aaMss <- cifreg(Event(time,status)~dnr.f0+dnr.f1+preauto+ttt24+cluster(id),data=wdata,weights=wdata$weights,cause=2)
summary(aaMss)
ll <- mediatorSurv(aaMss,fit,data=dat,wdata=wdata)
summary(ll)
if (runb>0) { bll <- BootmediatorSurv(aaMss,fit,data=dat,k.boot=k.boot); summary(bll)}

### binomial outcome  ############################
aaMss <- binreg(Event(ftime,status)~dnr.f0+dnr.f1+preauto+ttt24+cluster(id),data=wdata,time=50,weights=wdata$weights,cens.weights=1,cause=2)
summary(aaMss)
ll <- mediatorSurv(aaMss,fit,data=dat,wdata=wdata)
summary(ll)
if (runb>0) { bll <- BootmediatorSurv(aaMss,fit,data=dat,k.boot=k.boot); summary(bll)}
```


Multinomial regression
======================

Also works with mediator with more than two levels

 * meditor: wmi in 4 categories
 * exposure: age in 4 categories

```{r, cache=TRUE, eval=fullVignette}
data(tTRACE)
dcut(tTRACE) <- ~. 

weightmodel <- fit <- mlogit(wmicat.4 ~agecat.4+vf+chf,data=tTRACE,family=binomial)
wdata <- medweight(fit,data=tTRACE)

aaMss <- binreg(Event(time,status)~agecat.40+ agecat.41+ vf+chf+cluster(id),data=wdata,time=7,weights=wdata$weights,cause=9)
summary(aaMss)
MultMed <- mediatorSurv(aaMss,fit,data=tTRACE,wdata=wdata)
```

```{r  results="hide", echo=FALSE }
## To save time building the vignettes on CRAN, we cache time consuming computations
if (fullVignette) {
  saveRDS(MultMed, "MultMed.rds")
} else {
  MultMed <- readRDS("MultMed.rds")
}
```

```{r}
summary(MultMed)
```

SessionInfo
============


```{r}
sessionInfo()
```
