---
# YAML header created by ox-ravel
title: Analysis of multivariate survival data
author: Klaus Holst & Thomas Scheike
---


# Overview

When looking at multivariate survival data with the aim of learning about the
dependence that is present, possibly after correcting for some covariates
different approaches are available in the mets package

-   Binary models and adjust for censoring with inverse probabilty of  censoring weighting
    -   biprobit  model
-   Bivariate surival models of Clayton-Oakes type
    -   With regression structure on dependence parameter
    -   With additive gamma distributed random effects
    -   Special functionality for polygenic random effects modelling
        such as ACE, ADE ,AE and so forth.
-   Plackett OR model model
    -   With regression structure on OR dependence parameter
-   Cluster stratified Cox

Typically it can be hard or impossible to specify random effects models with special
structure among the parameters of the random effects. This is possible for
our specification of the random effects models.

To be concrete about the model structure assume that we have paired binomial
data \( T_1, \delta_1, T_2, \delta_2,  X_1, X_2 \) where the censored
survival responses are \( T_1, \delta_1, T_2, \delta_2 \) and we
have covariates \( X_1, X_2 \).

The basic models assumes that each subject has a marginal on Cox-form
\[
\lambda_{s(k,i)}(t) \exp( X_{ki}^T \beta)
\]
where \(s(k,i)\) is a strata variable.


## Gamma distributed frailties

The focus of this vignette is describe how to work on bivariate survival data using the
addtive gamma-random effects models. We present two different ways of specifying
different dependence structures.

-   Univariate models with a single random effect for each cluster and with
    a regression design on the variance.
-   Multivariate models with multiple random effects for each cluster.

The univariate models are
then given a given cluster random effects \(Z_k\) with
parameter \(\theta\) the joint survival function is given by the Clayton copula
and on the form
\[ \psi(\theta, \psi^{-1}(\theta,S_1(t,X_{k1}) ) + \psi^{-1}(\theta, S_1(t,X_{k1}) )
\]
where \(  \psi \) is the Laplace transform of a gamma distributed random
variable with mean 1 and variance \(\theta\).

We then model the variance within clusters by a cluster specific
regression design  such that
\[
  \theta = h(z_j^T \alpha)
\]
where \(z\) is the regression design (specified by theta.des  in the software),
and \(h\) is link function, that is either \(exp\) or the identity.

This model  can be fitted using a pairwise likelihood or the pseudo-likelihood
using either

-   twostage
-   twostageMLE

 To make the twostage approach possible we need a model with specific structure for the
 marginals.  Therefore given the
random effect of the clusters the survival distributions within a cluster
are independent and on the form
\[
 P(T_j > t| X_j,Z) = exp( -Z \cdot \Psi^{-1}(\nu^{-1},S(t|X_j)) )
\]
with \(\Psi\) the laplace of the gamma distribution with mean 1 and variance \(1/\nu\).


## Additive Gamma frailties

For the multivariate models we are given a multivarite random effect each cluster
\(Z=(Z_1,...,Z_d) \) with d random effects.
The total random effect for each subject \(j\) in  a cluster is then specified using a
regression design on these random effects, with a regression vector
\( V_j \) such that the total random effect is
  \( V_j^T (Z_1,...,Z_d) \).  The elements of \(V_J\) are 1/0.
The
random effects \( (Z_1,...,Z_d) \) has associated parameters \( (\lambda_1,...,\lambda_d) \)
and  \( Z_j \) is Gamma distributed with

-   mean \( \lambda_j/V_1^T \lambda \)
-   variance \( \lambda_j/(V_1^T \lambda)^2 \)

The key assumption to make the two-stage fitting possible is that
\[
   \nu =V_j^T \lambda
\]
is constant within clusters.  The consequence of this is that
the total random effect for each subject within a cluster,
\( V_j^T (Z_1,...,Z_d) \) , is gamma distributed with variance \(1/\nu\).

The DEFAULT parametrization (var.par=1) uses the variances of the random effecs
\[
 \theta_j  = \lambda_j/\nu^2
\]
 For alternative parametrizations one can specify that the parameters are  \(\theta_j=\lambda_j\) with the argument var.par=0.

Finally the parameters \( (\theta_1,...,\theta_d) \) are related to the parameters
of the model by a regression construction \( M \) (d x k), that links the \( d  \)
\( \theta \)  parameters
with the \( k \) underlying \( \alpha  \) parameters
\[
 \theta = M  \alpha.
\]
The default is a diagonal matrix for \(M\).
This can be used to make structural assumptions about the variances of the random-effects
as is needed for the ACE model for example. In the software \( M \) is called theta.des

Assume that the marginal survival distribution for subject \(i\) within cluster \(k\) is given
by \(S_{X_{k,i}}(t)\) given covariates \(X_{k,i}\).

Now given the random effects of the cluster \(Z_k\) and the covariates\(X_{k,i}\) \(i=1,\dots,n_k\)
we assume that subjects within the cluster are independent with survival distributions

\begin{align*}
  \exp(-  ( V_{k,i} Z_k)  \Psi^{-1} (\nu,S_{X_{k,i}}(t)) ).
\end{align*}

A consequence of this is that the hazards given the covariates \(X_{k,i}\) and the random effects \(Z_k\)
are given by

\begin{align}
  \lambda_{k,i}(t;X_{k,i},Z_{k,i}) = ( V_{k,i} V_k) D_3 \Psi^{-1} (\nu,S_{X_{k,i}}(t))  D_t S_{X_{k,i}}(t)
  \label{eq-cond-haz}
\end{align}

where \(D_t\) and \(D_3\) denotes the partial derivatives with respect to \(t\) and the third argument, respectively.

Further, we can express the multivariate survival distribution as

\begin{align}
  S(t_1,\dots,t_m) & =  \exp( -\sum_{i=1}^m (V_i Z) \Psi^{-1}(\eta_l,\nu_l,S_{X_{k,i}}(t_i)) )  \nonumber \\
  & =  \prod_{l=1}^p  \Psi(\eta_l,\eta , \sum_{i=1}^m Q_{k,i} \Psi^{-1}(\eta,\eta,S_{X_{k,i}}(t_i))).
  \label{eq-multivariate-surv}
\end{align}

In the case of considering just pairs, we write this function as \(C(S_{k,i}(t),S_{k,j}(t))\).

In addition to survival times from this model, we assume that we independent right censoring present
\(U_{k,i}\) such that the given \(V_k\) and the covariates\(X_{k,i}\) \(i=1,\dots,n_k\) \((U_{k,1},\dots,U_{k,n_k})\)
of \((T_{k,1},\dots,T_{k,n_k})\), and the conditional censoring distribution do not depend on \(V_k\).

One consequence of the model strucure is that the Kendall's can be computed for
two-subjects \((i,j)\) across two clusters \`\`1'' and \`\`2'' as

\begin{align}
E( \frac{( V_{1i} Z_1-  V_{1j}Z_2)( V_{2i}Z_1 -  V_{2j}Z_2 )}{( V_{1i}Z_1 + V_{2i}Z_2 ) ( V_{1j}Z_1 + V_{2j}Z_2 )} )
\end{align}

under the assumption that that we compare pairs with equivalent marginals,
\(S_{X_{1,i}}(t)= S_{X_{2,i}}(t)\) and \(S_{X_{1,j}}(t)= S_{X_{2,j}}(t)\),
and that \(S_{X_{1,i}}(\infty)= S_{X_{1,j}}(\infty)=0\).
Here we also use that \(\eta\) is the same across clusters.
The Kendall's tau would be the same for \eqref{frailty-model} due to the same additive structure for the
frailty terms, and the random effects thus have the same interpretation in terms of Kendall's tau.


## Univariate gamma (clayton-oakes) model twostage models

We start by fitting simple Clayton-Oakes models for the data, that is
with an overall random effect that is Gamma distrubuted with variance
\( \theta \). We can fit the model by a pseudo-MLE (twostageMLE)
and a pairwise composite likelihood approach (twostage).

The pseudo-liklihood and the composite pairwise likelhood should give
the same for this model since we have paired data.
In addition the log-parametrization is illustrated with the
var.link=1 option.
In addition it is specified that we want a "clayton.oakes" model.
We note that the standard errors differs because the twostage does not
include the variance due to the baseline parameters for this type of
modelling, so here it is better to use the twostageMLE.

```{r  twostage1 }
library(mets)
data(diabetes)

# Marginal Cox model  with treat as covariate
margph <- phreg(Surv(time,status)~treat+cluster(id),data=diabetes)
# Clayton-Oakes, MLE
fitco1<-twostageMLE(margph,data=diabetes,theta=1.0)
summary(fitco1)

# Clayton-Oakes
fitco2 <- survival.twostage(margph,data=diabetes,theta=0.0,detail=0,
                 clusters=diabetes$id,var.link=1,model="clayton.oakes")
summary(fitco2)
fitco3 <- survival.twostage(margph,data=diabetes,theta=1.0,detail=0,
                 clusters=diabetes$id,var.link=0,model="clayton.oakes")
summary(fitco3)
```

The marginal models can be either structured Cox model or as here
with a baseline for each strata. This gives quite similar results to those
before.

```{r  twostagemarginal }
# without covariates but marginal model stratified
marg <- phreg(Surv(time,status)~+strata(treat)+cluster(id),data=diabetes)
fitcoa <- survival.twostage(marg,data=diabetes,theta=1.0,clusters=diabetes$id,
		 model="clayton.oakes")
summary(fitcoa)
```


## Piecewise constant Clayton-Oakes model

Let the cross-hazard ratio (CHR) be defined as

\begin{align}
  \eta(t_1,t_2) =  \frac{ \lambda_1(t_1| T_2=t_2)}{ \lambda_1(t_1| T_2 \ge t_2)}
  =  \frac{ \lambda_2(t_2| T_1=t_1)}{ \lambda_2(t_2| T_1 \ge t_1)}
\end{align}

where \(\lambda_1\) and \(\lambda_2\) are the conditional hazard functions
of \(T_1\) and \(T_2\) given covariates.  For the Clayton-Oakes model this
ratio is \(\eta(t_1,t_2) = 1+\theta\), and as a consequence we see that
if the co-twin is dead at any time we would increase our risk
assessment on the hazard scale with the constant \(\eta(t_1,t_2)\). The
Clayton-Oakes model also has the nice property that Kendall's tau is
linked directly to the dependence parameter \(\theta\) and is \(1/(1+2/\theta)\).

A very useful extension of the model the constant
cross-hazard ratio (CHR) model is the
piecewise constant cross-hazard ratio (CHR) for bivariate survival
data [@nan2006piecewise], and this model was extended to competing
risks in [@shih2010modeling].

In the survival setting we let the CHR

\begin{align}
  \eta(t_1,t_2) & = \sum \eta_{i,j} I(t_1 \in I_i, t_2 \in I_j)
\end{align}

The model lets the CHR by constant in different part of the plane. This can be thought of also
as having a separate
Clayton-Oakes model for each of the regions specified in the plane here by
the cut-points \( c(0,0.5,2) \) thus defining 9 regions.

This provides a constructive goodness of fit test for the whether the Clayton-Oakes model
is valid. Indeed if valid the parameter should be the same in all regions.

First we generate some data from the Clayton-Oakes model with variance \(0.5\) and 2000 pairs.
And fit the related model.

```{r  claytonoakes }
d <- simClaytonOakes(2000,2,0.5,0,3)
 margph <- phreg(Surv(time,status)~x+cluster(cluster),data=d)
# Clayton-Oakes, MLE
fitco1<-twostageMLE(margph,data=d)
summary(fitco1)
```

 Now we cut the region at the cut-points
\( c(0,0.5,2) \) thus defining 9 regions and fit a separate model for each region.
We see that the parameter is indeed rather constant over the 9 regions. A formal
test can be constructed.

```{r  piecewise }
udp <- piecewise.twostage(c(0,0.5,2),data=d,score.method="optimize",
                          id="cluster",timevar="time",
                          status="status",model="clayton.oakes",silent=0)
summary(udp)
```


## Multivariate gamma twostage models

To illustrate how the multivariate models can be used, we first set
up some twin data with ACE structure. That is two shared random effects,
one being the genes \(\sigma_g^2\) and one the environmental effect
\( \sigma_e^2 \). Monozygotic twins share all genes whereas the dizygotic
twins only share half the genes. This can be expressed via 5 random effect for
each twin pair (for example). We start by setting this up.

 The pardes matrix tells how the the parameters of the 5 random effects are
 related, and the matrix her first has one random effect with parameter
 \(\theta_1\) (here the \( \sigma_g^2 \) ), then the next 3 random effects have
 parameters \(0.5 \theta_1\) (here \( 0.5 \sigma_g^2 \) ), and the last
random effect that is given by its own parameter \(\theta_2\) (here \(\sigma_e^2\) ).

```{r  twindesign }
data <- simClaytonOakes.twin.ace(2000,2,1,0,3)

out <- twin.polygen.design(data,id="cluster")
pardes <- out$pardes
pardes
```

The last part of the model structure is to decide how the random effects
are shared for the different pairs (MZ and DZ), this is specfied by
the random effects design (\(V_1\) and \(V_2\)) for each pair.
This is here specified by an overall designmatrix for each subject
(since they enter all pairs with the same random effects design).

For an MZ pair the two share the full gene random effect and the full
environmental random effect. In contrast the DZ pairs share the
2nd random effect with half the gene-variance and have both a non-shared
gene-random effect with half the variance, and finally a fully shared
environmental random effect.

```{r  desrv }
des.rv <- out$des.rv
# MZ
head(des.rv,2)
# DZ
tail(des.rv,2)
```

Now we call the twostage function. We see that we essentially recover the true values, and
note that the output also compares the sizes of the genetic and environmental random effect.
This number is sometimes called the heritability.  In addition the total variance for each
subject is also computed and is here around \(3\), as we indeed constructed.

```{r  twostage2 }
aa <- phreg(Surv(time,status)~x+cluster(cluster),data=data)
ts <- twostage(aa,data=data,clusters=data$cluster,detail=0,
 	      theta=c(2,1),var.link=0,step=0.5,
 	      random.design=des.rv,theta.des=pardes)
summary(ts)
```

The estimates can be transformed into Kendall's tau estimates for MZ and DZ twins. The Kendall's tau
in the above output reflects how  a gamma distributed random effect in the normal Clayton-Oakes model
is related to the Kendall's tau. In this setting the Kendall's of MZ and DZ, however, should reflect
both random effects.

We do this based on simulations. The Kendall's tau of the MZ is around 0.60, and for
DZ around 0.33. Both are quite high and this is due to a large shared environmental
effect and large genetic effect.

```{r  kendall }
kendall.ClaytonOakes.twin.ace(ts$theta[1],ts$theta[2],K=10000)
```


## Family data

For family data, things are quite similar since we use only the pairwise structure.
We show  how the designs are specified.

First we simulate data from an ACE model.  2000 families with two-parents that share
only the environment, and two-children that share genes with their parents.

```{r  familysim }
library(mets)
set.seed(1000)
data <- simClaytonOakes.family.ace(2000,2,1,0,3)
head(data)
data$number <- c(1,2,3,4)
data$child <- 1*(data$number==3)
```

To set up the random effects some functions can be used. We here set up the ACE model
that has 9 random effects with one shared environmental effect (the last random effect) and
4 genetic random effects for each parent, with variance \(\sigma_g^2/4\).

The random effect is again set-up with an overall designmatrix because it is again the same for
each subject for all comparisons across family members.  We below demonstrate how the model can
be specified in various other ways.

Each child share 2 genetic random effects with each parent, and also share 2 genetic random effects
with his/her sibling.

```{r  familydes }
out <- ace.family.design(data,member="type",id="cluster")
out$pardes
head(out$des.rv,4)
```

Then we fit the model

```{r  familytwostage }
pa <- phreg(Surv(time,status)~+1+cluster(cluster),data=data)
aa <- aalen(Surv(time,status)~+1,data=data,robust=0)

# make ace random effects design
ts <- twostage(pa,data=data,clusters=data$cluster,
	var.par=1,var.link=0,theta=c(2,1),
        random.design=out$des.rv,theta.des=out$pardes)
summary(ts)
```

The model can also be fitted by specifying the pairs that
one wants for the pairwise likelhood. This is done by
specifying the pairs argument.
We start by considering all pairs as we also did before.

All pairs can be written up by calling the familycluster.index
function.

There are 12000 pairs to consider and the last 12 pairs for the last family
is written out here.

```{r  familypairs }
# now specify fitting via specific pairs
# first all pairs
mm <- familycluster.index(data$cluster)
head(mm$familypairindex,n=10)
pairs <- matrix(mm$familypairindex,ncol=2,byrow=TRUE)
tail(pairs,n=12)
```

Then fitting the model using only specified pairs

```{r  familypairfit }
ts <- twostage(pa,data=data,clusters=data$cluster,
               theta=c(2,1),var.link=0,step=1.0,
               random.design=out$des.rv,
               theta.des=out$pardes,pairs=pairs)
summary(ts)
```

Now we only use a random sample of the pairs by sampling these.  The pairs
picked still refers to the data given in the data argument, and clusters (families)
are also specified as before.

```{r  familysample }
ssid <- sort(sample(1:12000,2000))
tsd <- twostage(aa,data=data,clusters=data$cluster,
               theta=c(2,1)/10,var.link=0,step=1.0,
               random.design=out$des.rv,
	      theta.des=out$pardes,pairs=pairs[ssid,])
summary(tsd)
```

Sometimes one only has the data from the pairs in addition to for example
a cohort estimate of the marginal surival models. We now demonstrate how this
is dealt with. Everything is essentially as before but need to organize the
design differently compared to before we specified the design  for everybody in
the cohort.

```{r  fmailycohortdes }
ids <- sort(unique(c(pairs[ssid,])))

pairsids <- c(pairs[ssid,])
pair.new <- matrix(fast.approx(ids,c(pairs[ssid,])),ncol=2)
head(pair.new)

# this requires that pair.new refers to id's in dataid (survival, status and so forth)
# random.design and theta.des are constructed to be the array 3 dims via individual specfication from ace.family.design
dataid <- dsort(data[ids,],"cluster")
outid <- ace.family.design(dataid,member="type",id="cluster")
outid$pardes
head(outid$des.rv)
```

Now fitting the model using only the pair data.

```{r  fmailycohortdesest }
tsdid <- twostage(aa,data=dataid,clusters=dataid$cluster,
               theta=c(2,1)/10,var.link=0,step=1.0,
               random.design=outid$des.rv,
               theta.des=outid$pardes,pairs=pair.new)
summary(tsdid)
coef(tsdid)
coef(tsd)
```

Now we illustrate how one can also
directly specify the  random.design and theta.design
for each pair,
rather than taking the rows of the des.rv for the relevant pairs.
This can be much simpler in some situations.

```{r  familyrandomdes }
pair.types <-  matrix(dataid[c(t(pair.new)),"type"],byrow=T,ncol=2)
head(pair.new)
head(pair.types)

# here makes pairwise design , simpler random.design og pardes, parameters
# stil varg, varc
# mother, child, share half rvm=c(1,1,0) rvc=c(1,0,1),
# thetadesmcf=rbind(c(0.5,0),c(0.5,0),c(0.5,0),c(0,1))
#
# father, child, share half rvf=c(1,1,0) rvc=c(1,0,1),
# thetadescf=rbind(c(0.5,0),c(0.5,0),c(0.5,0),c(0,1))
#
# child, child,  share half rvc=c(1,1,0) rvc=c(1,0,1),
# thetadesmf=rbind(c(0.5,0),c(0.5,0),c(0.5,0),c(0,1))
#
# mother, father, share 0 rvm=c(1,0) rvf=c(0,1),
# thetadesmf=rbind(c(1,0),c(1,0),c(0,1))

theta.des  <- array(0,c(4,2,nrow(pair.new)))
random.des <- array(0,c(2,4,nrow(pair.new)))
# random variables in each pair
rvs <- c()
for (i in 1:nrow(pair.new))
{
	if (pair.types[i,1]=="mother" & pair.types[i,2]=="father")
	{
	theta.des[,,i] <- rbind(c(1,0),c(1,0),c(0,1),c(0,0))
       	random.des[,,i] <- rbind(c(1,0,1,0),c(0,1,1,0))
	rvs <- c(rvs,3)
	} else {
  	theta.des[,,i] <- rbind(c(0.5,0),c(0.5,0),c(0.5,0),c(0,1))
	random.des[,,i] <- rbind(c(1,1,0,1),c(1,0,1,1))
	rvs <- c(rvs,4)
	}
}
# 3 rvs here
random.des[,,7]
theta.des[,,7]
# 4 rvs here
random.des[,,1]
theta.des[,,1]
head(rvs)
```

And fitting again the same model as before

```{r  familydesestimate }
tsdid2 <- twostage(aa,data=dataid,clusters=dataid$cluster,
               theta=c(2,1)/10,var.link=0,step=1.0,
               random.design=random.des,
               theta.des=theta.des,pairs=pair.new,pairs.rvs=rvs)
summary(tsdid2)
tsd$theta
tsdid2$theta
tsdid$theta
```

Finally the same model structure can be setup based on a Kinship coefficient.

```{r  familykinship }
# simpler specification via kinship coefficient for each pair
kinship  <- c()
for (i in 1:nrow(pair.new))
{
if (pair.types[i,1]=="mother" & pair.types[i,2]=="father") pk1 <- 0 else pk1 <- 0.5
kinship <- c(kinship,pk1)
}
head(kinship,n=10)

out <- make.pairwise.design(pair.new,kinship,type="ace")
names(out)
# 4 rvs here , here independence since shared component has variance 0 !
out$random.des[,,9]
out$theta.des[,,9]
```

Same same

```{r  twostagekinship }
tsdid3 <- twostage(aa,data=dataid,clusters=dataid$cluster,
               theta=c(2,1)/10,var.link=0,step=1.0,
               random.design=out$random.design,
               theta.des=out$theta.des,pairs=pair.new,pairs.rvs=out$ant.rvs)
summary(tsdid3)
coef(tsdid3)
```


## Univariate plackett model twostage models

The copula known as the Plackett distribution, see [@plackett1965; @anderson1992time; @ghosh2006sjs], is on the form

\begin{align}
  C(u,v; \theta) =
  \begin{cases}
    \frac{ S - (S^2 - 4 u v \theta (\theta-a))}{2 (\theta -1)} & \mbox{ if } \theta \ne 1 \\
    u v & \mbox{ if } \theta = 1
  \end{cases}
\end{align}

with \(S=1+(\theta-1) (u + v)\).  With marginals \(S_i\) we now define the
bivariate survival function as \(C(u_1,u_2)=H(S_1(t_1),S_2(t_2))\) with
\(u_i=S_i(t_i)\).

The dependence parameter \(\theta\) has the nice interpretation that the
it is equivalent to the odds-ratio of all \(2 \times 2\) tables for
surviving past any cut of the plane \((t_1,t_2)\), that is
\[
\theta = \frac{ P(T_1 > t_1 | T_2 >t_2) P(T_1 \leq t_1 | T_2>t_2) }{P(T_1 > t_1 | T_2 \leq t_2) P(T_1 \leq t_1 | T_2 \leq t_2 ) }.
\]

One additional nice feature of the odds-ratio measure it that it is
directly linked to the Spearman correlation, \(\rho\), that can be
computed as

\begin{align}
  \frac{\theta+1}{\theta -1} - \frac{2 \theta}{(\theta-1)^2} \log(\theta)
\end{align}

when \(\theta \ne 1\), if \(\theta=1\) then \(\rho=0\).

This model has a more free parameter than the Clayton-Oakes model.

```{r  twostageplackett }
library(mets)
data(diabetes)

# Marginal Cox model  with treat as covariate
margph <- phreg(Surv(time,status)~treat+cluster(id),data=diabetes)
# Clayton-Oakes, MLE
fitco1<-twostageMLE(margph,data=diabetes,theta=1.0)
summary(fitco1)

# Plackett model
mph <- phreg(Surv(time,status)~treat+cluster(id),data=diabetes)
fitp <- survival.twostage(mph,data=diabetes,theta=3.0,Nit=40,
               clusters=diabetes$id,var.link=1,model="plackett")
summary(fitp)

# without covariates but with stratafied
marg <- phreg(Surv(time,status)~+strata(treat)+cluster(id),data=diabetes)
fitpa <- survival.twostage(marg,data=diabetes,theta=1.0,
                clusters=diabetes$id,score.method="optimize")
summary(fitpa)

fitcoa <- survival.twostage(marg,data=diabetes,theta=1.0,clusters=diabetes$id,
                 model="clayton.oakes")
summary(fitcoa)
```

With a regression design

```{r  plackettreg }
mm <- model.matrix(~-1+factor(adult),diabetes)
fitp <- survival.twostage(mph,data=diabetes,theta=3.0,Nit=40,
               clusters=diabetes$id,var.link=1,model="plackett",
	       theta.des=mm)
summary(fitp)
```

```{r  piecewiseplackett }
# Piecewise constant cross hazards ratio modelling

d <- subset(simClaytonOakes(2000,2,0.5,0,stoptime=2,left=0),!truncated)
udp <- piecewise.twostage(c(0,0.5,2),data=d,score.method="optimize",
                          id="cluster",timevar="time",
                          status="status",model="plackett",silent=0)
summary(udp)
```
